---
title: 'Novel Coronavirus (COVID-19) Data Analysis'
author: "VM"
date: "2023-10-16"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Data Source

This data is taken from the "COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University" [git repo](https://github.com/CSSEGISandData/COVID-19)

Ensure you have installed the required libaries.

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(tidyr)
library(prophet)
library(data.table)
library(dplyr)
library(plotly)
library(hrbrthemes)
```

## Load Data

Data is directly loaded from the source git repo.  While running locally ensure internet access and the repo url is correct

```{r}
us_confirmed_df <- read_csv('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv') 
```

## Extract Required Data

Here I removed the columns with redundant information, because we use US only data it is not required to keep that as column, as combined key column has, county, state and country I used a required column and also transpose data to give observation for data in each row.


```{r}
data = us_confirmed_df[-c(1:10)]
df <- transpose(data)
rownames(df) <- colnames(data)
colnames(df) <- rownames(data)
names(df) <- lapply(df[1, ], as.character)
df <- df[-1, ]
head(df)
```

## Inspect for Wayne county


```{r}
df_wayne = df['Wayne, Michigan, US']
df_wayne$'Wayne, Michigan, US' = as.numeric(as.character(df_wayne$'Wayne, Michigan, US'))
summary(df_wayne)
```


```{r}
df_wayne$ds <- row.names(df)

df_wayne$ds <- as.Date(df_wayne$ds, format='%m/%d/%y')
df_wayne$y <- df_wayne$'Wayne, Michigan, US'
p <- df_wayne %>%
  ggplot( aes(x=ds, y=y)) +
    geom_area(fill="#69b3a2", alpha=0.5) +
    geom_line(color="#69b3a2") +
    ylab("Confirmed Cases") +
    theme_ipsum()

p <- ggplotly(p)
p
```

## Model using Prophet

```{r}
m <- prophet(df_wayne)
future <- make_future_dataframe(m, periods = 365)
tail(future)
forecast <- predict(m, future)
tail(forecast[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])
plot(m, forecast)
```
## Investigating Time Series

```{r}
prophet_plot_components(m, forecast)
```
